---
title: "Comparison of Electoral College Districts, 2010 vs 2020"
output:
  pdf_document: default
Author: Painschab, Matthew
Date: 10/7/2020
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

library(tidyverse)

headtohead <- read.csv("./derived_data/headtohead.csv")

```

Description of data set
-----------------------
This data is derived from two sources: 
1) the most recent weighted polling average across competitive states and electoral college districts for Nebraska and Maine from fivethirtyeight and 
2) a data set of the current number of electoral college votes in place today based on the 1990 census and another based on the projected 2020 census and projected reapportionment of congressional districts that will be in place for the 2022 congressional elections and 2024 presidential election.

General description of the question
-----------------------------------
2020 is an extreme in that the census occurs every 10 years and reapportionment
happens at that time as well. Therefore, a presidential election that falls in
census year provides the most distortion in that the the changes in population
over the last ten years have not been reflected in the current electoral
college. My hypothesis is that this will have an important impact on the
final outcome of the electoral college.

## 1. Analyze how the average population per electoral vote varies and how this
## has changed from 2010 to 2020.

1A. Mean population per electoral college vote is higher in states that Biden is
leading by both the 2010 census and 2020 census, by about 60,000 people per 
electoral vote in both scenarios.

```{r, echo=FALSE}

totals2 <- headtohead %>% 
  group_by(winner) %>% 
  summarize(mean_pop_10 = mean(poppervote_2010), 
            mean_pop_20 = mean(poppervote_2020))
totals2
```

1B. This figure demonstrates how each state population has shifted from 2010
to 2019, showing how the average population per electoral vote has changed 
since the last census. The black dot is the population per electoral vote in 
2010 and the yellow dot is the population per electoral vote in 2020 (before 
reapportionment that will take place in 2022). They are arranged in order of 
population per electoral vote in 2010. Most notably, there is a very large range
of population per electoral vote (from around 200,000 in the smallest states)
to around 700,000 in the largest states. This effectively demonstrates the outsize
importance that small states get in selecting our president compared to their
population.

As you can see, almost every state has had their population grow since 2010 (with a few exceptions). Clearly, as demonstrated by the blue (Biden) and red (Trump) lines, states that are currently polling for Biden have gained much more population per electoral vote
in the last ten years.

```{r, echo=FALSE}
options(scipen = 100000)

pop2019vote2010 <- headtohead %>% 
  mutate(newpop_oldvote = pop_2019/votes_2010) %>% 
  filter(votes_2010>1)

pop2019vote2010 %>%
  mutate(state = fct_reorder(state, poppervote_2010)) %>%
ggplot(aes(y = state)) + 
  geom_segment(aes(x = poppervote_2010, xend = newpop_oldvote, yend = state, 
                   color = winner)) +
  geom_point(aes(x = poppervote_2010)) +
  geom_point(aes(x = newpop_oldvote, color = "yellow")) +
  scale_color_manual(values = c("blue", "red", "yellow")) +
  labs(x = "Population Per Electoral Vote") +
  theme(axis.title.y=element_blank(), legend.position = "none")

```

## 2. Review how the electoral college vote change from 2010 to 2020 would have 
## looked based on the results of the 2016 election.

2A. Electoral College distribution 2016 election by number of states and/or 
districts with 2010 electoral college.

```{r, echo=FALSE}
ggplot(headtohead, aes(x = votes_2010, fill = trump_clinton)) + 
     geom_histogram(alpha = 0.5, binwidth = 1) + 
     scale_fill_manual(values = c("blue", "red")) +
     labs(x = "Electoral Votes 2010", y = "Number of States/Districts", 
          fill = "2016")
```

2B. Electoral College distribution 2016 election by number of states and/or 
districts with projected 2020 electoral college.

```{r, echo=FALSE}
ggplot(headtohead, aes(x = votes_2020, fill = trump_clinton)) + 
     geom_histogram(alpha = 0.5, binwidth = 1) + 
     scale_fill_manual(values = c("blue", "red")) +
     labs(x = "Electoral Votes 2020", y = "Number of States/Districts", 
          fill = "2016")
```

2C. Summary of how the 2016 election results with current electoral map verus
projected 2020 electoral map. 

```{r, echo=FALSE}
totals_2016 <- headtohead %>% 
  group_by(trump_clinton) %>% 
  summarize(elec_votes_10 = sum(votes_2010), 
            elec_votes_20 = sum(votes_2020))
totals_2016
```

Conlusion Section 2: Shifts of population across states from 2010 to 2020 would not
have greatly changed the outcome of the election. In fact, Trump would have won
by a slightly larger margin under the projected 2020 electoral college map.

## 3. Review Electoral College Votes for 2020 election By 2010 Distribution and 
## 2020 Expected Distribution

3A. Electoral College distribution 2010 by number of states and/or districts with 
currently projected winner by color.

```{r, echo=FALSE}
ggplot(headtohead, aes(x = votes_2010, fill = winner)) + 
  geom_histogram(alpha = 0.5, binwidth = 1) + 
  scale_fill_manual(values = c("blue", "red")) +
  labs(x = "Electoral Votes 2010", y = "Number of States/Districts")

```

3B. Electoral College distribution 2020 by number of states and/or districts with 
currently projected winner by color.

```{r, echo=FALSE}
ggplot(headtohead, aes(x = votes_2020, fill = winner)) + 
  geom_histogram(alpha = 0.5, binwidth = 1) + 
  scale_fill_manual(values = c("blue", "red")) +
  labs(x = "Electoral Votes 2020", y = "Number of States/Districts")
```

3C. Summary of how the election would look based on current polls 
with current electoral map verus projected 2020 electoral map. 

```{r, echo=FALSE}
totals <- headtohead %>% 
  group_by(winner) %>% 
  summarize(elec_votes_10 = sum(votes_2010), 
            elec_votes_20 = sum(votes_2020))
totals
```

Conlusions Section 3: The majority of states currently projected to go for Donald Trump 
are small states with at or below 10 electoral votes per state. 
The exception is Texas. This reflects the poor polling position that 
President Trump is currently in. The changes expected in 2020 would have very 
little effect, in fact again, would move the electoral college vote 
slightly towards President Trump.
